---
source: Rmd
title: "Introduction to ggplot2"
teaching: 30
exercises: 0
questions:
- "---?"
objectives:
- "---?"
- "---?"
keypoints:
- "---?"
- "---?"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Learning Objectives

By the end of this session, you should be able to:

- Use the **ggplot2** package to create your first plot
- Understand how to use layers in ggplot2
- Be familiar with a few different methods for visualising continuous and discrete variables
- Be able to create subplots (facets) with ggplot2
- Be aware of some strategies for dealing with overplotted data



# Introduction

In this session, we will show you some examples of how to build a plot with the **ggplot2** package. The ggplot2 package implements a "layered grammar of graphics" where data visualisation is divided into layers.

The layers that you will need to "build up" to form a plot include:

- The data you want to visualise (e.g. heights and weights).
- An aesthetic mapping to represent variables (e.g. map height to the y-axis, map weight to the x-axis).
- A geometric way to represent the information (e.g. display with a point).



# Load ggplot2

First, let's load ggplot2 into memory.

```{r}
library(tidyverse)  # contains ggplot2
```



# Import data

We specify the type of the categorical variables to stop them being interpreted as text or numbers. This will make plotting easier later.

```{r}
df <- read_csv('data/synthetic_data_clean.csv',
               col_types = cols(
                 chemo = col_factor(),
                 chronic_rrt = col_factor(),
                 metastatic = col_factor(),
                 radiotx = col_factor(),
                 medical = col_factor(),
                 vital_status = col_factor(),
                 sex = col_factor(),
                 elective_surgical = col_factor(),
                 system = col_factor(ordered=TRUE,
                                     levels=as.character(1:10))
               ))
```



# Generic structure of ggplot commands

It takes a while to get used to the structure, but it's worth the effort!

```{r, eval = FALSE}
ggplot(data = date_frame_name) +
  style_of_plot(mapping = aes(x = vector_name1, y = vector_name2)) +
  other_arguments
```



# Your first plot

A scatter plot is a useful way to represent two continuous variables, and spot any correlations between the variables. We'll plot CRP against neutrophil count.

```{r}
ggplot(data = df) + geom_point(mapping = aes(x = crp, y = neutrophil))
```

And here's the same command, written more succinctly:

```{r, eval=F}
ggplot(df) + geom_point(aes(crp, neutrophil))
```



# Building a plot up, layer by layer

Observe how the plot is built up layer by layer, starting with the data, and moving to the aesthetics and then geoms.

```{r}
ggplot(data = df) +
  aes(x = weight) +
  aes(y = height) +
  geom_point() +
  geom_jitter() +
  aes(colour = sex) +
  geom_smooth(method = "lm", se = F) +
  ggtitle("Height and Weight by Sex") +
  xlab("Weight (Kg)") +
  ylab("Height (cm)") +
  scale_colour_viridis_d(end=0.8, alpha = 0.5) + 
  theme_minimal() +
  theme(legend.title = element_blank()) 
```



# Histograms

Histograms are useful for displaying counts of continuous data. Here we use the argument `binwidth = 1` which tells the histogram to count the number of PaO2 samples in "bins" of 1. i.e. 0-1, 1-2, 2-3 kPa and so on...

The representation is very sensitive to the bind width. Try changing it and see what happens.

```{r}
ggplot(data = df) +
  aes(x = lactate_1hr) +
  geom_histogram (binwidth = 1, fill ="grey",col = "black") +
  ggtitle("Arterial lactate (hour 1 of admission)") +
  xlab("Lactate") +
  theme_minimal()
```



# Density Plots

Density plots are analogous to histograms. Rather than binning the data, everything is smoothed. Just like histograms, they are sensitive to how this data this smoothing occurs. This can be adjusted with the `bw` argument (which stands for "bandwidth").

```{r}
ggplot(data = df) +
  aes(x = lactate_1hr) +
  geom_density(fill ="grey",
               colour = "black") +
  ggtitle("Arterial PaO2") +
  ggtitle("Arterial lactate (hour 1 of admission)") +
  xlab("Lactate") +
  theme_minimal()
```


# Other ways of visualising continuous variables

We can use box and whisker plots to visualise the median, IQR and range of 1-hour lactates in patients who were alive or dead at discharge:

```{r}
ggplot(df) + geom_boxplot(aes(vital_status, lactate_1hr))
```

Or use a violin plot to get more of an insight into the how the lactates are distributed in those 2 groups. Remember that (like the density plots we looked at earlier) the violins are smoothed estimates. 

```{r}
ggplot(df) + geom_violin(aes(vital_status, lactate_1hr))
```



# A pipeline for exploratory analysis

We can chain together commands (using the pipe operator) from different bits of tidyverse to enable exploratory analysis:

```{r}
df %>% 
  filter(urea < 2) %>% 
  ggplot() +
  geom_point(aes(na, k))
```



# Faceting Plots

Sometimes you might wish to encode information across multiple plots. This is the role of faceting. Here, we visualise the distributions of sodium levels in male and female patients. 

```{r}
ggplot(df) +
  geom_density(aes(na)) +
  facet_wrap(~ sex)
```

And we can split things up further, now using both sex and mortality:

```{r}
ggplot(df) +
  geom_histogram(aes(lactate_1hr)) +
  facet_grid(sex ~ vital_status)
```

We chain together commands to plot histograms for all the continuous variables in a tibble:

```{r}
df %>%
  keep(is.numeric) %>%
  gather() %>% 
  ggplot() +
  geom_histogram(aes(value)) +
  facet_wrap(~ key, scales = 'free')
```

And do the same for bar plots of discrete variables:

```{r}
df %>% 
  keep(is.factor) %>% 
  gather() %>% 
  ggplot() +
  geom_bar(aes(value)) +
  coord_flip() +  # change to horizontal bars
  facet_wrap(~ key, scale s= 'free')
```



# Dealing with overplotting and scaling issues

Consider this graph:

```{r}
g <- ggplot(df, aes(crp, neutrophil)) + theme_minimal()
g + geom_point()  # overplotted
```

It's difficult to see exactly what is going on in the bottom-left corner of the plot, where the bulk of the points lie, because many opaque points are plotted on top of one another. We can try and fix this in a few ways:

```{r}
g + geom_point(alpha=0.1)  # make the points transparent
g + geom_bin2d()  # turn the points into a heatmap instead
```

But the bulk of the data are still clumped in the corner. Let's try to help the situation by altering the axes.

We could just clip the axes:

```{r}
neut_crp_scatter <- g + geom_point(alpha=0.3)
neut_crp_scatter + ylim(0, 30) + xlim(0, 400)
```

But some data are then obscured. What about log transforming the axes?

```{r}
# Turn off minor gridlines, which would otherwise be confusing
neut_crp_scatter <- neut_crp_scatter
  + theme(panel.grid.minor = element_blank())

neut_crp_scatter + scale_x_continuous(trans='log2',
                                      breaks=scales::pretty_breaks(10))
```

The log transform is a little extreme here. Now the bulk of the data are on the right side of the plot. Let's try a square root transform instead:

```{r}
neut_crp_scatter + 
  scale_x_continuous(trans='sqrt', breaks=scales::pretty_breaks(20)) + 
  scale_y_continuous(trans='sqrt', breaks=scales::pretty_breaks(20))
```