---
title: "Introduction"
author: "Steve Harris"
institute: "The DataSciBC"
date: "2019/09/17 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
long_df <- read_csv(
  "~/_data/esicm/longitudinal-data.csv",
  col_types = "?dddddddddddddddddcdddcdddddd")
demo_df <- read_csv("~/_data/esicm/demographic-data.csv")
```

class: inverse, center, middle

# Welcome

---

# Global Learning Objectives

- What does everyone want to get out of the day?
- Introduce a new way of handling data:
  - Understand the importance of reproducibility
  - Encourge the use of written code, rather than point and click
- Leave with enough knowledge to push further independently

---

# Session Priorities

- Ensure everyone can access the RStudio server
- Introduce the course data
- Briefly introduce files and directories

---

# Preliminaries

1. Get connected to the wifi
1. [Install Slack](https://slack.com/intl/en-gb/) or use the web interface
  - You will have an invite to the course channel
  - Introduce youselves on Slack
1. Check you can login to R-Studio Server
1. Pair programming: buddy up
1. Live coding: we will code along with you

**Green stickies** up when done. **Red stickies** up if stuck!

---

# Introducing the Course Data

- Critical Care Health Informatics Collaborative (CC-HIC)
- Aggregated EHR data
- 263 variables
  - 1d: Demographics, risk, outcomes
  - 2d: Treatment and physiology
- Anonymised sample of 1000 patients

---

# The Tables

- demographic_data.csv
  - information that **doesn't** change about the patient
  - "1d data"
  - demographics like sex, age, weight, height etc.
  - *1 row per patient*, and *1 column per variable*.
  
```{r}
demo_df
```

---

# The Tables

- longitudinal_data.csv
  - information that **does** change about the patient
  - "2d data"
  - C-reactive protein, ionotrope administration, ventilation settings etc.
  - *1 row per hour per patient*, and *1 column per variable*
  
```{r}
long_df
```

---

# Things to Highlight

- Not every cell will be filled with data.
- E.g. CRP (once daily measure)
- many cells with `NA` (the way `R` represents missing information)

For a full description of the data that exists inside CC-HIC see [here](../resources/static/data_spec.png)

---

# Your Data

- We encourage you to bring along your own data.
- Try out some of the exercises on your own data.
- Feel free to approach us during ESICM for advice on your data.

---

# Files and directories

- Live code in R Studio
- Showcase `file.choose()`