---
title: "Introduction"
author: "Ed Palmer & Steve Harris"
institute: "The DataSciBC"
date: "2019/09/17 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
synth_data <- read_csv("../../../_episodes_rmd/data/synthetic_data_clean.csv")
```

class: inverse, center, middle

# Welcome

---

# Global Learning Objectives

- Introduce a new way of handling data:
  - Understand the importance of reproducibility
  - Encourge the use of written code, rather than point and click
- Leave with enough knowledge to push further independently

---

# Session Priorities

- Ensure everyone can access RStudio Cloud
  - Can work from your machine if you prefer
- Introduce the course data

---

# Preliminaries

1. Get connected to the wifi
1. [Install Slack](https://slack.com/intl/en-gb/) or use the web interface
  - You will have an invite to the course channel
  - Introduce youselves on Slack
1. Check you can login to R-Studio Cloud
1. Pair programming: buddy up
1. Live coding: we will code along with you

**Green stickies** up when done. **Red stickies** up if stuck!

---

# Introducing the Course Data

- Critical Care Health Informatics Collaborative (CC-HIC)
- Aggregated EHR data
- 263 variables
  - Demographics, risk, outcomes, treatment and physiology
- Synthetically derrived sample of "5000 patients"
- Data looks real, but it isn't

---

# The Tables

- synthetic_data_clean.csv
  - *1 row per patient*, and *1 column per variable*.
  
```{r}
synth_data
```

# Things to Highlight

- We will be starting with a "dirty" version of this data
- Will learn how to clean it up today

---

# Your Data

- We encourage you to bring along your own data
- Try out some of the exercises on your own data
- Feel free to approach us during the next 2 days for advice on your data
