---
title: "Data Viz with ggplot2"
author: "Claire Black"
institute: "The DataSciBC"
date: "2016/09/16 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r include = FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%", warning = FALSE, message = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
    options$out.height <- "99%"
    options$fig.width <- 16
    options$fig.height <- 8
  }
  options
})
```

```{css, echo = FALSE}
/* custom.css */
.my_borders {
  padding-top: 50px;
  padding-bottom: 50px;
  padding-left: 20px;
  padding-right: 20px;
  margin: 20px;
  line-height: 2;
}
.my_line_height{ 
  line-height: 0.75; 
}
.left-code {
  color: #777;
  width: 49%;
  height: 92%;
  float: left;
}
.right-plot {
  width: 49%;
  float: right;
  padding-left: 1%;
}
```

# Learning Objectives

## By the end of this session, you should be able to use the **ggplot2** package to build:

- scatter plots
- histograms and density plots
- box plots
- facetted charts
- maps

---

# ggplot2 package

## The ggplot2 package implements a "layered grammar of graphics" where data visualization is divided into parameters: 

- the data you want to visualise, e.g. height and weight
- aesthetics to represent variables, e.g. map height to the y-axis, map weight to the x-axis
- geometric objects, e.g. display with a point
- a coordinate system, e.g. use a standard cartesian axis
- the ability to statistical transform variables

---

# Mapping Aesthetics
 
- This is how the data are "mapped" to a visual dimension
- We are most familiar with `x` and `y` mappings
- Others might include:
    - Size
    - Shape
    - Colour/Fill

---

# Load ***tidyverse***  which includes ***ggplot*** 

```{r}
library(tidyverse)
```

---

# Import the course data

### We are going to use the long_df.csv and demo_df.csv files used in the previous session to learn how to create the layers of a ggplot

---

## Data files

```{r, include = FALSE}
long_df <- read_csv(
  "~/_data/esicm/longitudinal-data.csv",
  col_types = "?dddddddddddddddddcdddcdddddd")
demo_df <- read_csv("~/_data/esicm/demographic-data.csv")
```

```{r, eval = FALSE}
long_df <- read_csv("./data/longitudinal-data.csv")
demo_df <- read_csv("./data/demographic-data.csv")
```

---

class: inverse, center, middle

# Scatter Plots

---

class: my_borders

.left-code[
```{r dot1, fig.show="hide"}
ggplot(data = demo_df)
```
]

.right-plot[
![](`r knitr::fig_chunk("dot1", "png")`)
]

---

class: my_borders

.left-code[
```{r dot2, fig.show="hide"}
ggplot(data = demo_df) + 
  aes(x = weight) +
  aes(y = height)
```
]

.right-plot[
![](`r knitr::fig_chunk("dot2", "png")`)
]

---

class: my_borders

.left-code[
```{r dot3, fig.show="hide"}
ggplot(data = demo_df) + 
  aes(x = weight) +
  aes(y = height) +
  geom_point()
```
]

.right-plot[
![](`r knitr::fig_chunk("dot3", "png")`)
]

---

class: my_borders

.left-code[
```{r dot4, fig.show="hide"}
ggplot(data = demo_df) + 
  aes(x = weight) +
  aes(y = height) +
  geom_point() +
  geom_jitter()
```
]

.right-plot[
![](`r knitr::fig_chunk("dot4", "png")`)
]

---

class: my_borders

.left-code[
```{r dot5, fig.show="hide"}
ggplot(data = demo_df) + 
  aes(x = weight) +
  aes(y = height) +
  geom_point() +
  geom_jitter() +
  aes(col = sex)
```
]

.right-plot[
![](`r knitr::fig_chunk("dot5", "png")`)
]

---

class: my_borders

.left-code[
```{r dot6, fig.show="hide"}
ggplot(data = demo_df) + 
  aes(x = weight) +
  aes(y = height) +
  geom_point() +
  geom_jitter() +
  aes(col = sex) +
  geom_smooth(method = "lm", se = F)
```
]

.right-plot[
![](`r knitr::fig_chunk("dot6", "png")`)
]

---

class: my_borders

.left-code[
```{r dot7, fig.show="hide"}
ggplot(data = demo_df) + 
  aes(x = weight) +
  aes(y = height) +
  geom_point() +
  geom_jitter() +
  aes(col = sex) +
  geom_smooth(method = "lm", se = F) +
  ggtitle("Height and Weight by Sex") +
  xlab("Weight (Kg)") +
  ylab("Height (cm)") 
```
]

.right-plot[
![](`r knitr::fig_chunk("dot7", "png")`)
]

---

class: my_borders

.left-code[
```{r dot8, fig.show="hide"}
ggplot(data = demo_df) + 
  aes(x = weight) +
  aes(y = height) +
  geom_point() +
  geom_jitter() +
  aes(col = sex) +
  geom_smooth(method = "lm", se = F) +
  ggtitle("Height and Weight by Sex") +
  xlab("Weight (Kg)") +
  ylab("Height (cm)") +
  scale_colour_viridis_d(end=0.8,alpha = 0.5)
```
]

.right-plot[
![](`r knitr::fig_chunk("dot8", "png")`)
]

---

class: my_borders

.left-code[
```{r dot9, fig.show="hide"}
ggplot(data = demo_df) + 
  aes(x = weight) +
  aes(y = height) +
  geom_point() +
  geom_jitter() +
  aes(col = sex) +
  geom_smooth(method = "lm", se = F) +
  ggtitle("Height and Weight by Sex") +
  xlab("Weight (Kg)") +
  ylab("Height (cm)") +
  scale_colour_viridis_d( end=0.8, alpha = 0.5) + 
  theme_minimal() +
  theme(legend.title = element_blank()) 
```
]

.right-plot[
![](`r knitr::fig_chunk("dot9", "png")`)
]

---

class: inverse, center, middle

# Histograms

---

class: my_borders

.left-code[
```{r hist1, fig.show="hide"}
ggplot(data = long_df) +
  aes(x = pao2_abg) 
```
]

.right-plot[
![](`r knitr::fig_chunk("hist1", "png")`)
]

---

class: my_borders

.left-code[
```{r hist2, fig.show="hide",warning=FALSE}
ggplot(data = long_df) +
  aes(x = pao2_abg) +
  geom_histogram () 
```
]

.right-plot[
![](`r knitr::fig_chunk("hist2", "png")`)
]

---

class: my_borders

.left-code[
```{r hist3, fig.show="hide", warning=FALSE}
ggplot(data = long_df) +
  aes(x = pao2_abg) +
  geom_histogram (binwidth = 1, fill ="grey", 
                  col = "black") 
  
```
]

.right-plot[
![](`r knitr::fig_chunk("hist3", "png")`)
]

---

class: my_borders

.left-code[
```{r hist4, fig.show="hide"}
ggplot(data = long_df) +
  aes(x = pao2_abg) +
  geom_histogram (binwidth = 1, fill ="grey", 
                  col = "black") +
  ggtitle("Arterial PaO2") +
  xlab("PaO2")
  
```
]

.right-plot[
![](`r knitr::fig_chunk("hist4", "png")`)
]

---

class: my_borders

.left-code[
```{r hist5, fig.show="hide"}
ggplot(data = long_df) +
  aes(x = pao2_abg) +
  geom_histogram (binwidth = 1, fill ="grey",
                  col = "black") +
  ggtitle("Arterial PaO2") +
  xlab("PaO2") +
  theme_minimal() 
  
```
]

.right-plot[
![](`r knitr::fig_chunk("hist5", "png")`)
]

---

class: inverse, center, middle

# Density Plots

---

class: my_borders

.left-code[
```{r dp1, fig.show="hide"}
ggplot(data = long_df) +
  aes(x = pao2_abg) 
```
]

.right-plot[
![](`r knitr::fig_chunk("dp1", "png")`)
]

---

class: my_borders

.left-code[
```{r dp2, fig.show="hide",warning=FALSE}
ggplot(data = long_df) +
  aes(x = pao2_abg) +
  geom_density () 
```
]

.right-plot[
![](`r knitr::fig_chunk("dp2", "png")`)
]

---

class: my_borders

.left-code[
```{r dp3, fig.show="hide", warning=FALSE}
ggplot(data = long_df) +
  aes(x = pao2_abg) +
  geom_density (fill ="grey", 
                  col = "black") 
  
```
]

.right-plot[
![](`r knitr::fig_chunk("dp3", "png")`)
]

---

class: my_borders

.left-code[
```{r dp4, fig.show="hide"}
ggplot(data = long_df) +
  aes(x = pao2_abg) +
  geom_density(binwidth = 1, fill ="grey", 
                  col = "black") +
  ggtitle("Arterial PaO2") +
  xlab("PaO2")
  
```
]

.right-plot[
![](`r knitr::fig_chunk("dp4", "png")`)
]

---

class: my_borders

.left-code[
```{r dp5, fig.show="hide"}
ggplot(data = long_df) +
  aes(x = pao2_abg) +
  geom_density (fill ="grey",
                  col = "black") +
  ggtitle("Arterial PaO2") +
  xlab("PaO2") +
  theme_minimal() 
  
```
]

.right-plot[
![](`r knitr::fig_chunk("dp5", "png")`)
]

---

class: inverse, center, middle

# Box and Whisker Plots

---

## Chemotheray and Neutrophil Count

This time need to join demographic and longitudinal dataframes together

```{r}
long_demo <- left_join(demo_df, long_df, by = "episode_id")
```

---

class: my_line_height

## long-demo

```{r}
names(long_demo)
```

---

## Data format

```{r}
str(long_demo$chemo)
str(long_demo$neutrophil)
```

Box plots need categorical data.
Change data format from numeric to categorical.

```{r}
long_demo <- long_demo %>% 
  mutate(chemo = as.factor(chemo))
```

```{r}
str(long_demo$chemo)
```

---

class: my_borders

.left-code[
```{r box1, fig.show="hide"}
ggplot(data = long_demo) +
  aes(x = chemo) +
  aes(y = neutrophil) 
```
]

.right-plot[
![](`r knitr::fig_chunk("box1", "png")`)
]

---

class: my_borders

.left-code[
```{r box3, fig.show="hide"}
ggplot(data = long_demo) +
  aes(x = chemo) +
  aes(y = neutrophil) +
  aes(colour = chemo) +
  geom_boxplot() 
```
]

.right-plot[
![](`r knitr::fig_chunk("box3", "png")`)
]

---

class: my_borders

.left-code[
```{r box4, fig.show="hide"}
ggplot(data = long_demo) +
  aes(x = chemo) +
  aes(y = neutrophil) +
  aes(colour = chemo) +
  geom_boxplot() +
  xlab("Chemotherapy")+
  ylab("Neutrophil") +
  ggtitle("Chemotherapy and Neutrophil Count") 
```
]

.right-plot[
![](`r knitr::fig_chunk("box4", "png")`)
]

---

class: my_borders

.left-code[
```{r box5, fig.show="hide"}
ggplot(data = long_demo) +
  aes(x = chemo) +
  aes(y = neutrophil) +
  aes(colour = chemo) +
  geom_boxplot() +
  xlab("Chemotherapy")+
  ylab("Neutrophil") +
  ggtitle("Chemotherapy and Neutrophil Count") +
  theme_light() +
  theme(legend.position = "none")
```
]

.right-plot[
![](`r knitr::fig_chunk("box5", "png")`)
]

---

class: inverse, center, middle

# Facetted Plots

---

class: my_borders

.left-code[
```{r f1, fig.show="hide"}
ggplot(data = long_demo) +
  aes(y = fluid_balance) +
  aes(x = pao2_abg)
```
]

.right-plot[
![](`r knitr::fig_chunk("f1", "png")`)
]

---

class: my_borders

.left-code[
```{r f2, fig.show="hide"}
ggplot(data = long_demo) +
  aes(y = fluid_balance) +
  aes(x = pao2_abg) +
  geom_point(alpha = 0.5)
```
]

.right-plot[
![](`r knitr::fig_chunk("f2", "png")`)
]

---

class: my_borders

.left-code[
```{r f3, fig.show="hide"}
ggplot(data = long_demo) +
  aes(y = fluid_balance) +
  aes(x = pao2_abg) +
  geom_point(alpha = 0.5) +
  aes(colour = as.factor(steroids)) 
```
]

.right-plot[
![](`r knitr::fig_chunk("f3", "png")`)
]

---

class: my_borders

.left-code[
```{r f4, fig.show="hide"}
ggplot(data = long_demo) +
  aes(y = fluid_balance) +
  aes(x = pao2_abg) +
  geom_point(alpha = 0.5) +
  aes(colour = as.factor(steroids)) +
  facet_wrap(~as.factor(steroids)) 
```
]

.right-plot[
![](`r knitr::fig_chunk("f4", "png")`)
]

---

class: my_borders

.left-code[
```{r f5, fig.show="hide"}
ggplot(data = long_demo) +
  aes(y = fluid_balance) +
  aes(x = pao2_abg) +
  geom_point(alpha = 0.5) +
  aes(colour = as.factor(steroids)) +
  facet_wrap(~as.factor(steroids)) +
  xlab("PaO2")+
  ylab("Daily Fluid balance (ml)") +
  ggtitle("PaO2, Steroids and Fluid Balance")
```
]

.right-plot[
![](`r knitr::fig_chunk("f5", "png")`)
]

---

class: my_borders

.left-code[
```{r f6, fig.show="hide"}
ggplot(data = long_demo) +
  aes(y = fluid_balance) +
  aes(x = pao2_abg) +
  geom_point(alpha = 0.5) +
  aes(colour = as.factor(steroids)) +
  facet_wrap(~as.factor(steroids)) +
  xlab("PaO2")+
  ylab("Daily Fluid balance (ml)") +
  ggtitle("PaO2, Steroids and Fluid Balance") +
  ylim(-5025,5025)+
  xlim(0,70)
```
]

.right-plot[
![](`r knitr::fig_chunk("f6", "png")`)
]

---

class: my_borders

.left-code[
```{r f7, fig.show="hide"}
ggplot(data = long_demo) +
  aes(y = fluid_balance) +
  aes(x = pao2_abg) +
  geom_point(alpha = 0.5) +
  aes(colour = as.factor(steroids)) +
  facet_wrap(~as.factor(steroids)) +
  xlab("PaO2")+
  ylab("Daily Fluid balance (ml)") +
  ggtitle("PaO2, Steroids and Fluid Balance") +
  ylim(-5025,5025)+
  xlim(0,70) +
  theme_light() +
  theme(legend.position = "none")
```
]

.right-plot[
![](`r knitr::fig_chunk("f7", "png")`)
]

---

class: my_borders

.left-code[
```{r f8, fig.show="hide"}
ggplot(data = long_demo) +
  aes(y = fluid_balance) +
  aes(x = pao2_abg) +
  geom_point(alpha = 0.5) +
  aes(colour = as.factor(steroids)) +
  facet_wrap(~as.factor(steroids)) +
  xlab("PaO2")+
  ylab("Daily Fluid balance (ml)") +
  ggtitle("PaO2, Steroids and Fluid Balance") +
  ylim(-5025,5025)+
  xlim(0,70) +
  theme_light() +
  theme(legend.position = "none")
```
]

.right-plot[
![](`r knitr::fig_chunk("f8", "png")`)
]

---

class: inverse, center, middle

# Chlorepleth

---

# Chlorepleth

```{r,  include = FALSE}
postcodes <- read_csv(file = "../ggpostcodes.csv")
```

## We are going to use a synthetic dataset as we don't share CCHIC postcodes

```{r, eval = FALSE}
postcodes <- read_csv(file = "./ggpostcodes.csv")
```

## Additionally going to use 

```{r}
library(maps)
library(RColorBrewer)
UK <- map_data(map = "world", region = "UK")
```

Slightly easier to use colour map

---

class: my_borders

.left-code[
```{r pc1, fig.show="hide"}
ggplot(data = UK) 
```
]

.right-plot[
![](`r knitr::fig_chunk("pc1", "png")`)
]

---

class: my_borders
.left-code[
```{r pc2, fig.show="hide"}
ggplot(data = UK) +
 aes(x = long, y = lat, group = group)
```
]

.right-plot[
![](`r knitr::fig_chunk("pc2", "png")`)
]

---

class: my_borders

.left-code[
```{r pc3, fig.show="hide"}
ggplot(data = UK) +
 aes(x = long, y = lat, group = group) + 
  geom_polygon(fill = "#004529", alpha = 0.75) 
```
]

.right-plot[
![](`r knitr::fig_chunk("pc3", "png")`)
]

---

class: my_borders

.left-code[
```{r pc4, fig.show="hide"}
ggplot(data = UK) +
 aes(x = long, y = lat, group = group) + 
  geom_polygon(fill = "#004529", alpha = 0.75) +
  geom_point(data = postcodes, 
             aes(y =  latitude, x = longitude, group = group),
             colour = "#f7fcb9", size = 0.5) 
```
]

.right-plot[
![](`r knitr::fig_chunk("pc4", "png")`)
]

---

class: my_borders

.left-code[
```{r pc5, fig.show="hide"}
ggplot(data = UK) +
 aes(x = long, y = lat, group = group) + 
  geom_polygon(fill = "#004529", alpha = 0.75) +
  geom_point(data = postcodes, 
             aes(y = latitude, x = longitude, 
                 group = group),
             colour = "#f7fcb9", size = 0.5) +
  coord_map() 
```
]

.right-plot[
![](`r knitr::fig_chunk("pc5", "png")`)
]

---

class: my_borders

.left-code[
```{r pc6, fig.show="hide"}
ggplot() +
  geom_polygon(
    data = UK,
    mapping = aes(
      x = long, y = lat, group = group),
      fill = "#004529", alpha = 0.75) +
  geom_point(
    data = postcodes, 
    mapping = aes(
      y =  latitude, x = longitude,
      group = group),
    colour = "#f7fcb9", size = 0.5) +
  coord_map() +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
```
]

.right-plot[
![](`r knitr::fig_chunk("pc6", "png")`)
]

---

class: inverse, center, middle

# Now your turn

---
